---
title: "MM2026Pred_F"
author: "Phillip DeGroot"
date: "2026-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Installing packages 
```{r}
library(BradleyTerry2)
library(dplyr)
library(ggplot2)
library(readxl)
```

```{r}
matchups = read.csv("2026_Potential_Matchups.csv")
```


#Get Kenpom data for 68 Tourney teams from https://www.formulabot.com/datasets/kenpom-march-madness
```{r}
kenpom_all = read_excel("kenpom_2000_2026.xlsx")

tourney_teams = c("Akron", "Alabama", "Arizona", "Arkansas", "BYU", "Cal Baptist",
  "Clemson", "Connecticut", "Duke", "Florida", "Furman", "Georgia", "Gonzaga",
  "Hawaii", "High Point", "Hofstra", "Houston", "Howard", "Idaho", "Illinois",
  "Iowa", "Iowa St.", "Kansas", "Kennesaw St.", "Kentucky", "LIU", "Lehigh",
  "Louisville", "McNeese", "Miami FL", "Miami OH", "Michigan", "Michigan St.",
  "Missouri", "North Dakota St.", "N.C. State", "Nebraska", "North Carolina",
  "Northern Iowa", "Ohio St.", "Penn", "Prairie View A&M", "Purdue", "Queens",
  "SMU", "Santa Clara", "Siena", "South Florida", "St. John's", "Saint Louis",
  "Saint Mary's", "TCU", "Tennessee", "Tennessee St.", "Texas", "Texas A&M",
  "Texas Tech", "Troy", "UCF", "UCLA", "UMBC", "Utah St.", "VCU", "Vanderbilt",
  "Villanova", "Virginia", "Wisconsin", "Wright St.")

kenpom = kenpom_all |>
  filter(Year == 2026) |>
  filter(TeamName %in% tourney_teams)

kenpom
```
#Checking for mismached names
```{r}
csv_teams = unique(c(matchups$HigherSeed, matchups$LowerSeed))
setdiff(csv_teams, kenpom$TeamName)
```

#Change Kenpom data so names match matchup csv
```{r}
name_map <- c(
  "Iowa St."           = "Iowa St",
  "Michigan St."       = "Michigan St",
  "Ohio St."           = "Ohio St",
  "Utah St."           = "Utah St",
  "Wright St."         = "Wright St",
  "N.C. State"         = "NC State",
  "McNeese"            = "McNeese St",
  "Tennessee St."      = "Tennessee St",
  "St. John's"         = "St John's",
  "Saint Louis"        = "St Louis",
  "Saint Mary's"       = "St Mary's CA",
  "North Dakota St."   = "N Dakota St",
  "Prairie View A&M"   = "Prairie View",
  "Queens"             = "Queens NC",
  "LIU"                = "LIU Brooklyn",
  "Kennesaw St."       = "Kennesaw"
)

for (kp_name in names(name_map)) {
  kenpom$TeamName[kenpom$TeamName == kp_name] <- name_map[kp_name]
}
```

#Verify for loop 
```{r}
setdiff(csv_teams, kenpom$TeamName)
```


#Plot of teams efficiency score
```{r}
ggplot(kenpom, aes(x = reorder(TeamName, AdjEM), y = AdjEM)) +
  geom_col(aes(fill = AdjEM)) + 
  labs(x = NULL, y = "Kenpom Adjusted Efficency Scores", title = "Kenpom AdjEM for Tourney Teams") 
```
#Creates a Strength Score that values "average" efficency. Takes from "trapizoid of excellence" idea that teams with a very high or low pace of play need to be "better" at their craft to see success in March Madness. 
```{r}
mean_tempo = mean(kenpom$AdjTempo)

z = function(x) (x - mean(x)) / sd(x)

kenpom$z_AdjEM = z(kenpom$AdjEM)
kenpom$z_AdjDE = z(-kenpom$AdjDE)
kenpom$z_AdjOE = z(kenpom$AdjOE)
kenpom$z_Tempo = z(-abs(kenpom$AdjTempo - mean_tempo))

#Creates a composite score that gives highest value towards effiency metrics 
kenpom$composite = 0.65 * kenpom$z_AdjEM + 0.15 * kenpom$z_AdjDE + 0.15 * kenpom$z_AdjOE + 0.05 * kenpom$z_Tempo

#Sanity Check

kenpom |>
  arrange(desc(composite))
```

#Find Sigma for Logistic Model  
```{r}
historical = kenpom_all |>
  filter(!is.na(seed)) |>
  filter(Year < 2026)

#finds historical difference in em for higher seeds

em_diffs = historical |>
  group_by(Year, seed) |>
  summarise(mean_em = mean(AdjEM), .groups = "drop") |>
  filter(seed <= 8) |>
  mutate(opp_seed = 17 - seed) |>
  left_join(
    historical |>
      group_by(Year, seed) |>
      summarise(opp_em = mean(AdjEM), .groups = "drop"), 
    by = c("Year", "opp_seed" = "seed")
  ) |>
  mutate(em_gap = mean_em - opp_em) |>
  group_by(seed) |>
  summarise(avg_em_diff = mean(em_gap), .groups = "drop") |>
  arrange(seed)

em_diffs


#historical win rates from higher seeds from https://www.fanduel.com/research/what-are-the-odds-of-a-perfect-march-madness-ncaa-tournament-bracket 

historical_winpc = c(
  0.988, # 1 vs 16
  0.931, # 2 vs 15
  0.856, # 3 vs 14
  0.794, # 4 vs 13
  0.644, # 5 vs 12
  0.613, # 6 vs 11
  0.613, # 7 vs 10
  0.481 # 8 vs 9
)

sigmas = seq(0, 25, by = 0.01)
sse = sapply(sigmas, function(s) {
  predicted = 1 / (1 + 10^(-em_diffs / s))
  sum((predicted - historical_winpc)^2)
})

sigma = as.numeric(unlist(sigmas[which.min(sse)]))

# Sanity check

sigma
```

#Fitting Logistic model to find win% against average 
```{r}
em_scale = sd(kenpom$AdjEM)
kenpom$em_equiv = kenpom$composite * em_scale
kenpom$p_avg = 1 / (1 + 10^(-kenpom$em_equiv / sigma))

kenpom |>
  arrange(desc(p_avg))
```

#Gets Log5 h2h probabilites 
```{r}
log5 = function(pA, pB) (pA- pA * pB) / (pA + pB - 2 * pA *pB)

teams = unique(kenpom$TeamName)
p_look = setNames(kenpom$p_avg, teams) #lookup function

pairs = expand.grid(t1 = teams, t2 = teams, stringsAsFactors = FALSE) |>
  filter(t1 < t2)
pairs$prob1 = log5(p_look[pairs$t1], p_look[pairs$t2])

set.seed(02222025)
n_sim = 1000
pairs$w1 = rbinom(nrow(pairs), n_sim, pairs$prob1)
pairs$w2 = n_sim - pairs$w1
```

#Bradley Terry Model
```{r}
teams_sorted = sort(teams)
bt = data.frame(
  player1 = factor(pairs$t1, levels = teams_sorted),
  player2 = factor(pairs$t2, levels = teams_sorted),
  wins1 = pairs$w1, wins2 = pairs$w2
)

bt_model = BTm(cbind(wins1, wins2), player1, player2, data = bt)

bt_model
```

```{r}
ability = BTabilities(bt_model)

adf = data.frame(
  team = rownames(ability), ability = ability[, "ability"], se = ability[, "s.e."], stringsAsFactors = FALSE
) 

adf |>
  arrange(desc(ability))
```

#Ability CI plot
```{r}
ggplot(adf, aes(x = reorder(team, ability), y = ability)) +
  geom_point() +
  geom_errorbar(aes(ymin = ability - 1.96 * se, ymax = ability + 1.96 * se)) +
  labs(X = NULL, y = "BT Ability", title = "BT Abilities of MM field")
```


#Predictions
```{r}
abl = setNames(adf$ability, adf$team)

matchups$Predictions = round(
  exp(abl[matchups$HigherSeed]) / 
    (exp(abl[matchups$HigherSeed]) + exp(abl[matchups$LowerSeed])),
  4
)

out = matchups |>
  select(HigherSeed, HigherSeedID, HigherSeedNum, LowerSeed, LowerSeedID, LowerSeedNum, Predictions)
write.csv(out, "Phil_D_2026_MM_Pred.csv", row.names = FALSE)
```

#Simulating the 2026 Bracket and picking the team with higher win%

```{r}
pick = function(a, b) {
  la = abl[a]; lb <- abl[b]
  pa = exp(la) / (exp(la) + exp(lb))
  if (pa >= 0.5) list(w = a, p = round(pa, 3)) else list(w = b, p = round(1 - pa, 3))
}

#First Four
ff = list(
  list("NC State", "Texas", "11E"),       
  list("SMU", "Miami OH", "11MW"),        
  list("Prairie View", "Lehigh", "16S"),   
  list("Howard", "UMBC", "16MW")           
)
ffw = list()
for (g in ff) {
  r = pick(g[[1]], g[[2]])
  ffw[[g[[3]]]] = r$w
  cat(sprintf("  %s vs %s -> %s (%.1f%%)\n", g[[1]], g[[2]], r$w, r$p * 100))
}
```

#Region Brackets
```{r}
East = c("Duke","Siena", "Ohio St","TCU",
           "St John's","Northern Iowa", "Kansas","Cal Baptist",
           "Louisville",ffw[["11E"]], "Michigan St","N Dakota St",
           "UCLA","UCF", "Connecticut","Furman")

South = c("Florida",ffw[["16S"]], "Clemson","Iowa",
            "Vanderbilt","McNeese St", "Nebraska","Troy",
            "North Carolina","VCU", "Illinois","Penn",
            "St Mary's CA","Texas A&M", "Houston","Idaho")

Midwest = c("Michigan",ffw[["16MW"]], "Georgia","St Louis",
              "Texas Tech","Akron", "Alabama","Hofstra",
              "Tennessee",ffw[["11MW"]], "Virginia","Kennesaw",
              "Kentucky","Santa Clara", "Iowa St","Tennessee St")

West = c("Arizona","LIU Brooklyn", "Villanova","Utah St",
           "Wisconsin","High Point", "Arkansas","Hawaii",
           "BYU","South Florida", "Gonzaga","Wright St",
           "Miami FL","Missouri", "Purdue","Queens NC")

```

#Region Sims
```{r}
sim_region = function(t16, rname) {

  cat("-- Round of 64 --\n")
  w64 = character(8)
  for (i in 1:8) {
    a = t16[2*i - 1]; b = t16[2*i]
    r = pick(a, b); w64[i] = r$w
    cat(sprintf("  %s vs %s -> %s (%.1f%%)\n", a, b, r$w, r$p * 100))
  }

  cat("-- Round of 32 --\n")
  w32 = character(4)
  for (i in 1:4) {
    r = pick(w64[2*i-1], w64[2*i]); w32[i] = r$w
    cat(sprintf("  %s vs %s -> %s (%.1f%%)\n", w64[2*i-1], w64[2*i], r$w, r$p*100))
  }

  cat("-- Sweet 16 --\n")
  w16 = character(2)
  for (i in 1:2) {
    r = pick(w32[2*i-1], w32[2*i]); w16[i] = r$w
    cat(sprintf("  %s vs %s -> %s (%.1f%%)\n", w32[2*i-1], w32[2*i], r$w, r$p*100))
  }

  cat("-- Elite Eight --\n")
  r = pick(w16[1], w16[2])
  cat(sprintf("  %s vs %s -> %s (%.1f%%)\n", w16[1], w16[2], r$w, r$p*100))
  return(r$w)
}

e_champ  = sim_region(East, "East")
s_champ  = sim_region(South, "South")
mw_champ = sim_region(Midwest, "Midwest")
w_champ  = sim_region(West, "West")
```

#Final Four
```{r}
cat("-- Semifinal 1: East vs South --\n")
r1 = pick(e_champ, s_champ)
cat(sprintf("  %s vs %s -> %s (%.1f%%)\n", e_champ, s_champ, r1$w, r1$p * 100))

cat("-- Semifinal 2: West vs Midwest --\n")
r2 = pick(w_champ, mw_champ)
cat(sprintf("  %s vs %s -> %s (%.1f%%)\n", w_champ, mw_champ, r2$w, r2$p * 100))

cat("\n========== NATIONAL CHAMPIONSHIP ==========\n")
rf = pick(r1$w, r2$w)
cat(sprintf("  %s vs %s -> %s WINS! (%.1f%%)\n", r1$w, r2$w, rf$w, rf$p * 100))
cat(sprintf("\n*** 2026 PREDICTED CHAMPION: %s ***\n", rf$w))
```




